---
title: "Democracy Around the World: A Data Analysis"
format: 
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    number-sections: true
embed-resources: true
code-fold: true
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(purrr)
library(countrycode)
library(ggridges)
library(scales)
library(gt)
```

```{r load-data, message=FALSE}
democracy_data <- readr::read_csv(
  'https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-11-05/democracy_data.csv'
)
masterSet <- read.csv("masterSet.csv")
```

# Data Context

## Original Data

For this study, we used a dataset from the Boix-Miller-Rosato Dichotomous Coding of Democracy dataset, which was shared through the TidyTuesday project on GitHub. The dataset provides a detailed look at political systems around the world from 1950 through 2022. Because each row represents a single country in a single year, the data allow us to examine political patterns and changes across more than seven decades of global history.

The dataset was created and curated by political scientists using historical records, expert coding, government documents, and electoral information. Its main purpose is to support research on topics such as democratization, regime stability, election integrity, suffrage, and political representation. This makes it especially useful for studying how democratic and non-democratic systems differ, as well as how political institutions evolve over time. Some of the most important variables include government type, democracy status, full suffrage, postponed elections, female political leadership, and electoral system type. These variables give us a broad view of each country’s political structure and make it possible to compare trends across countries, regions, and years.

## Supplemental GDP and Unemployment Data

To look at factors connected to a country’s economic performance, we added two outside sources to our democracy project. These sources were collected using the API-Ninjas public API and gave us information about GDP, GDP growth, and unemployment. Adding these economic variables helps connect the political side of the project to economic outcomes, which matters because our questions focus on how regime type and leadership may relate to a country’s overall performance.

The first source, Country_GDP.csv, focuses on national economic output. It came from the API-Ninjas GDP endpoint, which provides country-level economic information from sources such as the World Bank and IMF. Each observation represents one country in one year, so the main unit of analysis is country-year. This source includes variables such as country code, year, nominal GDP, and GDP growth rate. The years available are not the same for every country, but many countries have records across several decades and into the 2020s. We collected this information by using the country codes from our democracy records, sending those country codes through the API, and combining the returned results into one table. This source is useful because GDP and GDP growth give us a way to compare economic performance across political systems.

The second source, employ2.csv, focuses on unemployment. It came from the API-Ninjas unemployment endpoint. Like the GDP source, each observation represents one country in one year, so the unit of analysis is also country-year. This source includes variables such as country code, year, and unemployment rate. The available years differ by country because the API does not return the same amount of information for every country. We collected this information using the same general process as before: we looped through the country codes from our democracy records, requested unemployment information for each country, and combined the results into one table. This source matters because unemployment gives another way to measure economic conditions beyond just national output.

After collecting both economic sources, we joined them with the original democracy records using three digit country code and the year of the given observation. This created our combined working table, which brings together political variables and economic variables for the analysis. One limitation is that not every country-year had matching GDP or unemployment values, so some missing values remained after the join. This makes sense because the political records and the API results do not cover every country and year in the same exact way.

------------------------------------------------------------------------

# Research Questions

This report addresses two research questions:

> **Question 1 (Original Data):** How have regime types changed over time across different continents, and which countries have experienced the most political instability?

This question draws exclusively on the democracy dataset, examining regime change counts by country and cumulative changes over time by continent.

> **Question 2 (Supplemental Data):** What key factors — including democracy status, unemployment rate, and female leadership — are associated with a country's GDP?

This question leverages the joined master dataset combining democracy, GDP, and unemployment data, allowing us to model and visualize economic outcomes as a function of political and social variables.

------------------------------------------------------------------------

# Data Summaries & Visualizations

## Question 1: Regime Changes Over Time

### Summary Tables

#### Countries with the Most Regime Changes

The function below counts the number of times a country changed its regime type between consecutive years, then identifies the top 10 most politically unstable countries over the full dataset period.

```{r regime-change-function}
count_regime_changes <- function(country) {
  country_data <- democracy_data |>
    filter(country_name == country) |>
    arrange(year)

  regime <- country_data$regime_category_index
  changes <- regime != lag(regime)
  sum(changes, na.rm = TRUE)
}

results <- democracy_data |>
  distinct(country_name) |>
  mutate(n_changes = map_dbl(country_name, count_regime_changes))

results |>
  slice_max(n_changes, n = 10) |>
  rename(Country = country_name, `Number of Regime Changes` = n_changes) |>
  gt() |>
  tab_header(
    title = "Top 10 Most Politically Unstable Countries",
    subtitle = "Total regime changes, 1950–2022"
  ) |>
  tab_caption("Table 1. Top 10 countries by total number of regime changes (1950–2022).") |>
  tab_style(
    style = list(cell_fill(color = "#18bc9c"), cell_text(color = "white", weight = "bold")),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = Country)
  ) |>
  opt_stylize(style = 1)
```

#### Regime Change Counts by Continent

This table summarizes the total and average number of regime changes per continent across the full dataset period, giving a higher-level geographic view of political instability to complement the country-level table above.

```{r regime-continent-table}
democracy_data |>
  arrange(country_name, year) |>
  group_by(country_name) |>
  mutate(regime_changed = regime_category_index != lag(regime_category_index)) |>
  ungroup() |>
  mutate(continent = countrycode::countrycode(
    country_name,
    origin = "country.name",
    destination = "continent"
  )) |>
  filter(!is.na(continent)) |>
  group_by(continent, country_name) |>
  summarise(country_changes = sum(regime_changed, na.rm = TRUE), .groups = "drop") |>
  group_by(continent) |>
  summarise(
    `Number of Countries` = n(),
    `Total Regime Changes` = sum(country_changes),
    `Average Changes per Country` = round(mean(country_changes), 2),
    `Most Unstable Country` = country_name[which.max(country_changes)],
    `Max Changes (single country)` = max(country_changes),
    .groups = "drop"
  ) |>
  arrange(desc(`Total Regime Changes`)) |>
  rename(Continent = continent) |>
  gt() |>
  tab_header(
    title = "Regime Change Counts by Continent",
    subtitle = "1950–2022"
  ) |>
  tab_caption("Table 2. Regime change counts by continent, including total changes, average per country, and the most politically unstable country in each region.") |>
  fmt_number(columns = `Average Changes per Country`, decimals = 2) |>
  tab_style(
    style = list(cell_fill(color = "#18bc9c"), cell_text(color = "white", weight = "bold")),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = Continent)
  ) |>
  opt_stylize(style = 1)
```

### Visualizations

#### Top 10 Countries by Total Regime Changes

```{r top-countries-regime-plot}
#| fig-alt: "Horizontal bar chart showing the top 10 countries with the most total regime changes between 1950 and 2022. Each bar is colored by continent, allowing readers to see which regions contribute the most politically unstable countries."

results |>
  slice_max(n_changes, n = 10) |>
  mutate(continent = countrycode::countrycode(
    country_name, origin = "country.name", destination = "continent"
  )) |>
  ggplot(aes(x = n_changes, y = reorder(country_name, n_changes), fill = continent)) +
  geom_col() +
  scale_fill_manual(values = c(
    "Africa"   = "#E63946",
    "Americas" = "#F4A261",
    "Asia"     = "#2A9D8F",
    "Europe"   = "#457B9D",
    "Oceania"  = "#9B5DE5"
  )) +
  scale_x_continuous(breaks = scales::breaks_pretty()) +
  labs(
    title = "Top 10 Most Politically Unstable Countries",
    subtitle = "Total regime changes, 1950–2022",
    x = "Number of Regime Changes",
    y = NULL,
    fill = "Continent",
    caption = "Source: Boix-Miller-Rosato Democracy Dataset via TidyTuesday"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.major.y = element_blank()
  )
```

#### Cumulative Regime Changes per Continent Over Time

```{r regime-change-continent-plot}
#| fig-alt: "Area and line chart showing cumulative regime changes over time from 1950 to 2022, broken out by continent. Africa and Asia show the steepest growth in cumulative changes. Each continent is represented by a distinct color."

regime_change_counts <- democracy_data |>
  arrange(country_name, year) |>
  group_by(country_name) |>
  mutate(regime_changed = regime_category_index != lag(regime_category_index)) |>
  ungroup() |>
  mutate(continent = countrycode::countrycode(
    country_name,
    origin = "country.name",
    destination = "continent"
  )) |>
  group_by(continent, year) |>
  summarise(n_changes = sum(regime_changed, na.rm = TRUE), .groups = "drop") |>
  group_by(continent) |>
  mutate(cumulative_changes = cumsum(n_changes)) |>
  ungroup()

ggplot(regime_change_counts,
       aes(x = year, y = cumulative_changes, fill = continent)) +
  geom_area(alpha = 0.3, position = "identity") +
  geom_line(aes(color = continent), linewidth = 0.9) +
  scale_color_manual(values = c(
    "Africa"   = "#E63946",
    "Americas" = "#F4A261",
    "Asia"     = "#2A9D8F",
    "Europe"   = "#457B9D",
    "Oceania"  = "#9B5DE5"
  )) +
  scale_fill_manual(values = c(
    "Africa"   = "#E63946",
    "Americas" = "#F4A261",
    "Asia"     = "#2A9D8F",
    "Europe"   = "#457B9D",
    "Oceania"  = "#9B5DE5"
  )) +
  labs(
    x = "Year",
    y = "Cumulative Regime Changes",
    title = "Cumulative Regime Changes Over Time by Continent",
    subtitle = "1950–2022",
    color = "Continent",
    fill = "Continent",
    caption = "Source: Boix-Miller-Rosato Democracy Dataset via TidyTuesday"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )
```

------------------------------------------------------------------------

## Question 2: Factors Associated with Countries' GDP

### Summary Tables

#### Proportion of Years with a Female Leader by Regime Type and Election System

The table below summarizes the proportion of country-years with a female leader (monarch or president), grouped by a user-specified variable.

```{r female-leader-function}
sum_female_leader <- function(democ_data, columns, group) {
  female_leader_any <- democ_data |>
    rowwise() |>
    mutate(female_leader = any(c_across(all_of(columns)), na.rm = TRUE)) |>
    ungroup()

  prop_female <- female_leader_any |>
    group_by(across(all_of(group))) |>
    summarise(prop_female_leader = mean(female_leader, na.rm = TRUE),
              .groups = "drop")

  return(prop_female)
}

# By regime category
sum_female_leader(
  democracy_data,
  columns = c("is_female_monarch", "is_female_president"),
  group = "regime_category"
) |>
  filter(!is.na(regime_category)) |>
  arrange(desc(prop_female_leader)) |>
  slice_head(n = 25) |>
  rename(`Regime Category` = regime_category,
         `Proportion with Female Leader` = prop_female_leader) |>
  gt() |>
  tab_header(
    title = "Female Leadership by Regime Type",
    subtitle = "Top 25 regime categories by proportion of country-years with a female leader"
  ) |>
  tab_caption("Table 3. Top 25 regime categories by proportion of country-years with a female leader.") |>
  fmt_percent(columns = `Proportion with Female Leader`, decimals = 1) |>
  tab_style(
    style = list(cell_fill(color = "#18bc9c"), cell_text(color = "white", weight = "bold")),
    locations = cells_column_labels()
  ) |>
  opt_stylize(style = 1)
```

```{r female-leader-election}
# By election system
sum_female_leader(
  democracy_data,
  columns = c("is_female_monarch", "is_female_president"),
  group = "election_system"
) |>
  filter(!is.na(election_system)) |>
  arrange(desc(prop_female_leader)) |>
  slice_head(n = 22) |>
  rename(`Election System` = election_system,
         `Proportion with Female Leader` = prop_female_leader) |>
  gt() |>
  tab_header(
    title = "Female Leadership by Election System",
    subtitle = "Proportion of country-years with a female leader"
  ) |>
  tab_caption("Table 4. Proportion of country-years with a female leader by election system.") |>
  fmt_percent(columns = `Proportion with Female Leader`, decimals = 1) |>
  tab_style(
    style = list(cell_fill(color = "#18bc9c"), cell_text(color = "white", weight = "bold")),
    locations = cells_column_labels()
  ) |>
  opt_stylize(style = 1)
```

### Visualizations

#### GDP Growth: Democracies vs. Non-Democracies Over Time

```{r gdp-democracy-plot}
#| fig-alt: "Line chart showing average annual GDP growth rate from the 1960s to 2022, separately for democracies (blue) and non-democracies (red). Both groups show similar volatility with spikes around global economic crises."

aveGrowth <- masterSet |>
  group_by(year, is_democracy) |>
  summarise(mean_growth = mean(gdp_growth, na.rm = TRUE), .groups = "drop") |>
  filter(!is.na(mean_growth), !is.na(is_democracy))

ggplot(aveGrowth, aes(x = year, y = mean_growth, color = factor(is_democracy))) +
  geom_line(linewidth = 1) +
  scale_color_manual(
    values = c("FALSE" = "#E63946", "TRUE" = "#2A9D8F"),
    labels = c("FALSE" = "Non-Democracy", "TRUE" = "Democracy")
  ) +
  scale_y_continuous(labels = label_number(suffix = "%")) +
  labs(
    x = "Year",
    y = "Average GDP Growth Rate",
    title = "Mean GDP Growth Rate: Democracies vs. Non-Democracies",
    color = NULL,
    caption = "Source: API-Ninjas GDP Data joined with TidyTuesday Democracy Dataset"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "top"
  )
```

#### GDP Growth Distribution by Female Leadership Status (2015–2020)

```{r gdp-female-ridge-plot, warning=FALSE, message= FALSE}
#| fig-alt: "Ridge plot showing the distribution of GDP growth rates for countries with and without a female leader, for each year from 2015 to 2020. Both groups show broadly similar distributions, though countries with female leaders show slightly different spread in some years."

female_leader_any <- masterSet |>
  rowwise() |>
  mutate(
    female_leader = any(
      c_across(c(is_female_monarch, is_female_president)) == 1,
      na.rm = TRUE
    )
  ) |>
  ungroup() |>
  filter(!is.na(gdp_growth))

female_leader_any |>
  filter(year > 2014) |>
  mutate(
    female_leader = if_else(female_leader, "Female Leader", "No Female Leader"),
    year = factor(year)
  ) |>
  ggplot(aes(x = gdp_growth, y = year, fill = female_leader)) +
  geom_density_ridges(
    alpha = 0.65,
    color = "white",
    scale = 0.99,
    rel_min_height = 0.01
  ) +
  scale_x_continuous(
    limits = c(-15, 20),
    breaks = seq(-15, 20, 5),
    labels = label_number(suffix = "%")
  ) +
  scale_fill_manual(values = c(
    "No Female Leader" = "#457B9D",
    "Female Leader"    = "#F4A261"
  )) +
  labs(
    title = "GDP Growth Distributions by Female Leadership Status",
    subtitle = "Countries grouped by whether they had a female leader, 2015–2020",
    x = "GDP Growth Rate",
    y = NULL,
    fill = NULL,
    caption = "Source: API-Ninjas GDP Data joined with TidyTuesday Democracy Dataset"
  ) +
  theme_ridges(font_size = 13, grid = TRUE) +
  theme(
    legend.position = "top",
    legend.justification = "left",
    plot.title = element_text(face = "bold", size = 16)
  )
```

------------------------------------------------------------------------

# Data Model

To formally address Research Question 2, we used a linear regression model to examine how economic and political factors are associated with a country’s GDP. The response variable in the model was the log of nominal GDP, while the predictor variables included year, unemployment rate, democratic status, and an interaction between unemployment rate and democratic status. Including year allowed us to account for overall economic growth over time, while unemployment rate served as an economic indicator that may help explain differences in GDP. Democratic status was included to test whether democracies and non-democracies tend to differ in GDP, and the interaction term allowed us to examine whether the relationship between unemployment and GDP changes depending on whether a country is democratic. We used the log of GDP because nominal GDP is highly right-skewed, meaning that a small number of countries have extremely large GDP values compared to the rest of the dataset. Taking the log helps reduce this skew, stabilize the variation in the data, and make the model coefficients easier to interpret. We also excluded the year 2020 from the model because the COVID-19 pandemic created unusual economic disruptions that could distort the overall relationship between GDP, unemployment, and democracy.

```{r model-filter}
mastfilt <- masterSet |>
  dplyr::filter(
    !is.na(gdp_nominal),
    !is.na(unemployment_rate),
    !is.na(year),
    !is.na(is_democracy),
    year != 2020
  )
```

```{r model-fit, warning=FALSE,message=FALSE,  results='hide'}
lmGDP <- lm(log(gdp_nominal) ~ unemployment_rate * is_democracy + year,
            data = mastfilt)
```

```{r model-table}
broom::tidy(lmGDP) |>
  mutate(
    term = case_when(
      term == "(Intercept)"                        ~ "Intercept",
      term == "unemployment_rate"                  ~ "Unemployment Rate",
      term == "is_democracyTRUE"                   ~ "Is Democracy",
      term == "year"                               ~ "Year",
      term == "unemployment_rate:is_democracyTRUE" ~ "Unemployment × Democracy"
    ),
    across(c(estimate, std.error, statistic), ~ round(.x, 4)),
    p.value = case_when(
      p.value < 0.001 ~ "< 0.001",
      TRUE ~ as.character(round(p.value, 3))
    )
  ) |>
  rename(
    Term = term,
    Estimate = estimate,
    `Std. Error` = std.error,
    `t-statistic` = statistic,
    `p-value` = p.value
  ) |>
  gt() |>
  tab_header(
    title = "Linear Regression: Predictors of Log(Nominal GDP)"
  ) |>
  tab_caption("Table 5. Linear regression model predicting log(Nominal GDP). Coefficients represent the change in log(GDP) for a one-unit increase in each predictor.") |>
  tab_style(
    style = list(cell_fill(color = "#18bc9c"), cell_text(color = "white", weight = "bold")),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_fill(color = "#f0faf7"),
    locations = cells_body(rows = c(3, 5))
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = Term)
  ) |>
  opt_stylize(style = 1)
```

**Key findings from the model:**

- **Year** has a significant positive coefficient, confirming that GDP has grown over time on average across all countries, regardless of political regime.
- **Unemployment rate** is negatively associated with log(GDP), as expected — higher unemployment is associated with smaller economies.
- **Is Democracy** is positively associated with log(GDP), suggesting democratic countries tend to have higher GDPs on average.
- The **interaction term** (Unemployment × Democracy) is negative and significant, meaning the negative relationship between unemployment and GDP is *stronger* in democracies than in non-democracies. In other words, rising unemployment is more damaging to GDP in democratic countries.

```{r model-plot, warning= FALSE, message= FALSE}
#| fig-alt: "Scatter plot with regression lines showing the relationship between unemployment rate and log GDP, separately for democracies (teal) and non-democracies (red). The regression line for democracies has a steeper negative slope, indicating unemployment is more negatively associated with GDP in democratic countries."

mastfilt <- mastfilt |>
  mutate(preds = exp(predict(lmGDP)))

ggplot(
  aes(y = log(gdp_nominal), x = unemployment_rate, color = factor(is_democracy)),
  data = mastfilt
) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  scale_color_manual(
    values = c("FALSE" = "#E63946", "TRUE" = "#2A9D8F"),
    labels = c("FALSE" = "Non-Democracy", "TRUE" = "Democracy")
  ) +
  labs(
    x = "Unemployment Rate (%)",
    y = "Log(Nominal GDP)",
    color = "Regime Type",
    title = "Relationship Between Unemployment and Log GDP by Democracy Status",
    caption = "Source: API-Ninjas GDP & Unemployment Data joined with TidyTuesday Democracy Dataset"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "top"
  )
```
