---
title: "BradenProject2Check4"
format: html
editor: 
  markdown: 
    wrap: 72
---

## What are some key factors that are associated with countries' GDPs?

We filtered our data-set to where it has no missing values for gdp,
unemployment,whether they are a democracy or not and year. We omiited
the year 2020 to the dataset because this year skewed the data due to
the pandemic.

```{r}
masterSet<-read.csv("masterSet.csv")

mastfilt <- masterSet |>
  dplyr::filter(!is.na(gdp_nominal),
                !is.na(unemployment_rate),
                !is.na(year),
                !is.na(is_democracy),
                year != 2020)
```

We created a model to predict a function of GDP (log(GDP)) from the year
as well as the interaction of unemployment rate and whether that country
was a democracy or not. We had to use a function of GDP because there
was abnormalities in the GDP that complicate the model.

\[ \log(\text{GDP}\_i) = \beta\_0 + \beta\_1 ,\text{Unemployment}\_i +
\beta\_2 ,\text{Democracy}\_i + \beta\_3 ,(\text{Unemployment}\_i
\times \text{Democracy}\_i) + \beta\_4 ,\text{Year}\_i + \varepsilon\_i
\]

```{r}
lmGDP = lm(log(gdp_nominal) ~ unemployment_rate*is_democracy + year, data = mastfilt)
summary(lmGDP)
```

```{r}
mastfilt <- mastfilt |>
  mutate(
    preds = exp(predict(lmGDP)))
```
Here we are adding the predicted GDP values to the original dataset. 

```{r}
ggplot(aes(y =log(gdp_nominal), x = unemployment_rate, color = factor(is_democracy)), data = mastfilt) +
  geom_smooth(method = 'lm', se = F) +
  geom_point(alpha = 0.1) +
  labs(
    x = "Unemployment Rate",
    y = "Log GDP",
    color = "Democracy Status",
    title = "Relationship Between Unemployment and Log GDP by Democracy Status") +
  scale_color_discrete(
    labels = c("Non-Democracy", "Democracy"))
```
This plot shows the relationship between the interaction of whether a country is democratic or not and the unemployment rate on our function of GDP. We notice a faster drop in our function GDP when a country is democratic and the unemployment rate increases compared to when it is not democratic. 
```{r}
ggplot(data = mastfilt, aes(x = year, y = preds)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Year",
    y = "Predicted GDP",
    title = "Predicted GDP Over Time") +
  theme_minimal()
```
This plot demonstrates the relationship between the year and GDP. As the plot demonstrates for every year we notice as increase in GDP which is what we would expect.


