---
title: "Project 1 Checkpoint 3"
format: html
embed-resources: true
editor: source
execute: 
  echo: true
  error: true
---


```{r}
#| label: packages-setup
#| message: false
#| warning: false

library(httr)
library(tidyjson)
library(jsonlite)
library(tidyverse)
library(glue)
```

##Gathering GDP Data
I used the API listed for the Country GDP over a set number of years, I built a function to iterate over all of the country 3 letter alpha codes to pull all of their GDP data. After I had all of the GDP data for the countries in my democracy dataset, I used a bind rows by country_code as the id to create one large data frame.

```{r}
democracy_data <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-11-05/democracy_data.csv')
# Importing the data

country_code <- democracy_data |> 
    distinct(
        country_code
        ) |> 
    pull(country_code)
# Extracting the Country Codes 

 usa_df <- democracy_data |>
    filter(country_code == "USA")
 # Taking only USA

```

```{r}
get_gdp <- function(country, key = my_key){
    
    url <- glue("https://api.api-ninjas.com/v1/gdp?country={country};X-Api-Key={key}")
    # My Url for API GDP data
    
    my_data <- safely(fromJSON)(url)
    # Safely to grab data regardless errors
    
    if (is.null(my_data$result)) {
        return(NA)
    } else {
        return(my_data$result)
    }
    # Procaution against missing Countries. 
    
}
get_gdp("USA")
# Test 

country_gdp <- map(country_code, get_gdp) 
# All data for each distinct country in my Democracy data set

```

```{r}
country_gdp_df <- country_gdp |>
    setNames(country_code
             ) |> 
    bind_rows(
        .id = "country_code"
    )
```

##Gathing Unemployment Datas
I used the API listed for the Unemployment Data over a set number of years, I built a function to iterate over all of the country 3 letter alpha codes to pull all of their GDP data. After I had all of the Unemployement data for the countries in my democracy dataset, I used a bind rows by country_code as the id to create one large data frame.

```{r}
country = "CAN"
baseurl = "https://api.api-ninjas.com/v1/unemployment?country="
can_url = glue("{baseurl}{country};X-Api-Key={key}")

can_data = fromJSON(can_url)
```

```{r}
funky_Employ <- function(country_code){
  baseurl = "https://api.api-ninjas.com/v1/unemployment?country="
  url = glue("{baseurl}{country_code};X-Api-Key={key}")
  employ_data = fromJSON(url)
  return(employ_data)
}
can_data <- funky_Employ(country)
```

```{r}
democracy_data <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-11-05/democracy_data.csv')
democ_countrycode <- democracy_data |>
  select(country_code) |>
  distinct(country_code) |>
  pull(country_code)
```

```{r}
employ <- map(democ_countrycode, funky_Employ)
#employ2 <- map_dfr(democ_countrycode, funky_Employ)
write.csv(employ2, "employ2.csv", row.names = FALSE)
employ <- read.csv("employ2.csv")
#4543 rows
employ |> 
  distinct(country_code) |>
  count()
#110 countries in this 

names(employ) <- c("country_code", "year", "unemployment_rate")

#check for duplicate countries & years
democracy_data |>
  count(country_code, year) |>
  filter(n > 1)

employ |>
  add_count(country_code, year) |>
  filter(n > 1)
#shows that there are 40 country code duplicates
employ<- employ |>
  distinct()

employ_democracy_data <- democracy_data |>
  left_join(employ, by = c("country_code", "year"))

employ_democracy_data |>
  summarize(missing_unemployment = sum(is.na(unemployment_rate)))
#11195 missing values which makes sense because this api only has data for 110 countries and limited years

write.csv(employ_democracy_data, "employ_democracy_data.csv", row.names = FALSE)

```
##All the Data Combind
```{r}
# Democracy data
democracy_data<-readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-11-05/democracy_data.csv')
# GDP data
gdpData <- read.csv("Country_GDP.csv")

# Employment data
employData<-read.csv("employ2.csv")

#Combining the two data sets found through API's
combined <- left_join(
  gdpData,
  employData,
  by = c("year", "country"),
  relationship = "many-to-many"
)
# Combining with orginal democracy_data 
masterSet<-left_join(democracy_data, combined, by=c('country_code','year'))

write_csv(masterSet,file="masterSet.csv")

```

```{r}
masterSet<-read.csv("masterSet.csv")
```

##Visualizing the relationship between GDP growth and if a country is a democracy, over time. 

```{r}
aveGrowth <- masterSet %>%
  group_by(year, is_democracy) %>%
  summarise(mean_growth = mean(gdp_growth, na.rm = TRUE), .groups = "drop") %>% 
  filter(!is.na(mean_growth), !is.na(is_democracy))
  
ggplot(aveGrowth, aes(x=year,y=mean_growth,color=is_democracy))+
  geom_line(linewidth = 1)+
  labs(x="Year",
       y="Average Growth in GDP (%)",
       title="Mean Growth in GDP for Democracies vs. Non-Democracies",
       color="Is Democracy")+
  scale_color_manual(
  values = c("TRUE" = "blue", "FALSE" = "red"),
  labels = c("FALSE" = "Non-democracy", "TRUE" = "Democracy")
)
  
```

##Visualizing the change in unemployment rate over time for countries that are communist vs. are not communist. 

```{r}
aveRate<- masterSet %>% 
  group_by(year, is_communist) %>% 
  summarise(mean_rate = mean(unemployment_rate,na.rm=TRUE, .groups='drop')) %>% 
  filter(!is.na(mean_rate), !is.na(year))

ggplot(aveRate, aes(x=year,y=mean_rate,color=is_communist))+
  geom_line(linewidth=1)+
  labs(x ="Year",y ="Average Unemployment Rate (%)", title = "Mean Change in Unemployment Rate for Communist vs Non-Communist Countries")+
  scale_color_manual(
  values = c("TRUE" = "blue", "FALSE" = "red"),
  labels = c("FALSE" = "Not Communist", "TRUE" = "Communist")
)

```
