---
title: "Project1 Check 4 Final"
format: html
embed-resources: true
code-fold: true
---

```{r,message=FALSE}
library(countrycode)
library(tidyverse)
library(ggridges)
library(scales)
```

```{r, message=FALSE}
democracy_data <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-11-05/democracy_data.csv')

```

```{r}
count_regime_changes_per_year <- function(yr) {
  democracy_data |>
    arrange(country_name, year) |>
    group_by(country_name) |>
    mutate(regime_changed = regime_category_index != lag(regime_category_index)) |>
    ungroup() |>
    filter(year == yr) |>
    summarise(regime_changes = sum(regime_changed, na.rm = TRUE)) |>
    pull(regime_changes)
}

regime_change_counts <- democracy_data |>
  arrange(country_name, year) |>
  group_by(country_name) |>
  mutate(regime_changed = regime_category_index != lag(regime_category_index)) |>
  ungroup() |>
  mutate(continent = countrycode::countrycode(country_name,
                                              origin = "country.name",
                                              destination = "continent")) |>
  group_by(continent, year) |>
  summarise(n_changes = sum(regime_changed, na.rm = TRUE), .groups = "drop") |>
  group_by(continent) |>
  mutate(cumulative_changes = cumsum(n_changes)) |>
  ungroup()



```

## Visualizing regime changes per continent over time ##
```{r}
#| fig-alt: "Graph visualizing the number of regime changes over time per continent"

ggplot(regime_change_counts, mapping = aes(x=year,y = cumulative_changes,
                                           fill = continent))+
  geom_area(alpha = 0.3, position = "identity") +
  geom_line(aes(color = continent))+
  labs(x = "Year", y= "Cumulative Changes",
       title = "Number of Regime Changes over Time per Continent ")+
  theme_minimal()+
  scale_fill_manual(values = c(
  "Africa"   = "#E63946",
  "Americas" = "#F4A261",
  "Asia"     = "#2A9D8F",
  "Europe"   = "#457B9D",
  "Oceania"  = "#9B5DE5"
))
  
```

```{r}
masterSet<-read.csv("masterSet.csv")


sum_female_leader <- function(democ_data, columns, group){ 
    
    female_leader_any <- democ_data |> 
        rowwise() |>    # Selects the rows as groups to check over each year
        mutate(female_leader = any(c_across(all_of(columns)), na.rm = TRUE)
               ) |>
                        # Checks across defined columns for a female leader
        ungroup()       # Undoes rowwise() grouping
    
    prop_female <- female_leader_any |> 
        group_by(across(all_of(group))) |>  # Grouped by group = argument
        summarise(prop_female_leader = mean(female_leader, na.rm = TRUE))
                
                        # mean of TRUE/FALSE for the proportion 

    return(prop_female) # returns the group argument and the proportion
}

female_lead <- sum_female_leader(masterSet, 
                     columns = c("is_female_monarch", "is_female_president"), 
                     group = "country_name") |>
    arrange(-prop_female_leader)

```

\
```{r}
female_leader_any <- masterSet |>
  rowwise() |>
  mutate(
    female_leader = any(c_across(c(is_female_monarch, is_female_president)) == 1, na.rm = TRUE)
  ) |>
  ungroup()


female_leader_any <- female_leader_any |> filter(!is.na(gdp_growth)) 

female_leader_any |> 
    count(female_leader, year) |>
    filter(female_leader == TRUE) |>
    arrange(-n)


female_leader_any |>
    filter(year > 2014) |>
    mutate(
    female_leader = if_else(female_leader, "Female leader", "No female leader"),
    year = factor(year)
    ) |> 
ggplot(mapping = aes(x = gdp_growth, y = as.factor(year), fill = female_leader)) +
    geom_density_ridges(alpha = 0.65,
                        color = "white",
                        scale = 0.99,
                        rel_min_height = 0.01) +
    scale_x_continuous(limits = c(-15,20), 
                       breaks = seq(-15,20,5),
                       labels = label_number(suffix = "%")) +
    scale_fill_manual(
        values = c(
        "No female leader" = "steelblue",
        "Female leader" = "hotpink"
    )) +
    labs(title = "GDP Growth Distributions by Female Leadership Status",
         subtitle = "Countries grouped by whether they had a female leader, 2015 to 2020",
         x = "GDP growth rate",
         y = NULL,
         fill = NULL) + 
    theme_ridges(font_size = 13,
                 grid = TRUE) +
    theme(
        legend.position = "top",
        legend.justification = "left",
        plot.title = element_text(face = "bold", size = 18),
        plot.caption = element_text(size = 10, color = "gray40")
    )

```


## What are some key factors that are associated with countries' GDPs?

We filtered our data-set to where it has no missing values for gdp,
unemployment,whether they are a democracy or not and year. We omiited
the year 2020 to the dataset because this year skewed the data due to
the pandemic.

```{r}
masterSet<-read.csv("masterSet.csv")

mastfilt <- masterSet |>
  dplyr::filter(!is.na(gdp_nominal),
                !is.na(unemployment_rate),
                !is.na(year),
                !is.na(is_democracy),
                year != 2020)
```

We created a model to predict a function of GDP (log(GDP)) from the year
as well as the interaction of unemployment rate and whether that country
was a democracy or not. We had to use a function of GDP because there
was abnormalities in the GDP that complicate the model.

```{r}
lmGDP = lm(log(gdp_nominal) ~ unemployment_rate*is_democracy + year, data = mastfilt)
summary(lmGDP)
```

```{r}
mastfilt <- mastfilt |>
  mutate(
    preds = exp(predict(lmGDP)))
```
Here we are adding the predicted GDP values to the original dataset. 

```{r}
ggplot(aes(y =log(gdp_nominal), x = unemployment_rate, color = factor(is_democracy)), data = mastfilt) +
  geom_smooth(method = 'lm', se = F) +
  geom_point(alpha = 0.1) +
  labs(
    x = "Unemployment Rate",
    y = "Log GDP",
    color = "Democracy Status",
    title = "Relationship Between Unemployment and Log GDP by Democracy Status") +
  scale_color_discrete(
    labels = c("Non-Democracy", "Democracy"))
```
This plot shows the relationship between the interaction of whether a country is democratic or not and the unemployment rate on our function of GDP. We notice a faster drop in our function GDP when a country is democratic and the unemployment rate increases compared to when it is not democratic. 
```{r}
ggplot(data = mastfilt, aes(x = year, y = preds)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Year",
    y = "Predicted GDP",
    title = "Predicted GDP Over Time") +
  theme_minimal()
```
This plot demonstrates the relationship between the year and GDP. As the plot demonstrates for every year we notice as increase in GDP which is what we would expect.


